# Cleaning the Raw Data

Now that the database is built, the raw cougar and black bear data that I'm getting from Utah's Wildlife Tracker needs to be transformed into a format that can be imported into my monitoring techniques database. The data from Wildlife Tracker will go into my carnivore, capture_sites, collar, deployments, and gps_data tables. 

## Preliminary steps

### Load Libraries

Load all the libraries I'll need to run the code

```{r pkgs, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}

library(DBI)
library(RSQLite)
library(tidyverse)
library(lubridate)
library(sf)

```

### Database connection

Then make sure that there's still a connection with the monitoring techniques
database.

```{r db_connection2, eval = FALSE, echo = TRUE}

montech_db <- dbConnect(RSQLite::SQLite(),
                        "montech.db")

```

### Import the raw data

Import the raw data from wildlife tracker

```{r data, eval = FALSE, echo = TRUE}

raw_data <- read.csv("raw_data/2023-04-11_GPS_Data.csv", header = TRUE)

```

Once these preliminary steps are completed, we can actually clean the data.

## Formating Data for the Tables

### Carnivore table formating

I'll start with the data needed for the carnivores table. I'll need to pull the 
columns uniqueID, earTag, sex, birthYear, currentCohort, and species from the raw 
data using the `select()` function. Since these columns have different names than 
the names of my fields in my database table I then need to rename each column 
using the `rename()` function. Then, because there are many entries for each 
carnivore_id, I need to pull only the unique carnivore ids. In this case I used `filter(!duplicated(carnivore_id))` instead of distinct because Wildlife Tracker 
calculates the age class based on the birth year so there were duplicate rows for 
individuals that were unique because the collar had been on them for multiple age 
classes. Finally, I need to make sure that my entries matched the table constraints. 
This meant deleting all the spaces in the entries using the `gsub()` function.

```{r carnivore_data, eval = FALSE, echo = TRUE}

carnivores <- raw_data %>%
  select(uniqueID, earTag, sex, birthYear, currentCohort, species) %>%
  rename(carnivore_id = uniqueID, 
         ear_tag = earTag, 
         birth_year = birthYear,
         age_class = currentCohort) %>%
  arrange(age_class) %>%
  filter(!duplicated(carnivore_id)) %>%
  mutate(species = gsub(" ", "", species))

```

### Collar table formating

I'll next create the collars table. Again, I'll need to pull only the necessary
columns using the `select()` function. In this case, it's only collarID and freq.
Since Wildlife Tracker doesn't contain information on brand, model, or status I
had to create placeholder columns populated with NAs until I can acquire that data
from a different source. This time the columns are in a different order than
what I need for my database, so I used the `relocate()`function to reorder them.
I then renamed the necessary columns and finally pulled out the unique entries 
using the `distinct()` function.

```{r collar_data, eval = FALSE, echo = TRUE}

collars <- raw_data %>%
  select(collarID, freq) %>%
  mutate(brand = NA, model = NA, status = NA) %>%
  relocate(freq, .after = status) %>%
  rename(collar_id = collarID, 
         frequency = freq) %>%
  distinct()

```

### Deployments table formating

For the deployments table, I needed to pull carnivoreID and collarID from the 
raw Wildlife Tracker data. Since I don't yet have data on the deployment start
and end dates, I just created empty columns for those fields and populated them
with NA until I acquire the necessary data. Finally, I reordered the columns
according to my database format using `relocate()`and grabbed only the unique 
combinations of columns using `disctinct()`.

```{r deployments1_table, eval = FALSE, echo = TRUE}

deployments <- raw_data %>%
  select(uniqueID, collarID) %>%
  rename(carnivore_id = uniqueID, collar_id = collarID) %>%
  relocate(collar_id, .before = carnivore_id) %>%
  mutate(start_deployment = NA, end_deployment = NA) %>%
  distinct()

```

### Capture sites table formating 

For the capture sites table, I needed to pull the captureUnit, captureSubUnit,
and realCaptureArea columns from the Wildlife Tracker data using `select()`. I
then renamed them and switched the order of the columns so that it matched my
database table. As there were multiple columns with spaces in the entries, I had
to eliminate those using `gsub` so that it matched the database table constraints.
Lastly, I took only the unique combinations of entries by using `distinct()`.

```{r capture_sites_data, eval = FALSE, echo = TRUE}

capture_sites <- raw_data %>%
  select(captureUnit, captureSubUnit, realCaptureArea) %>%
  rename(capture_unit = captureUnit, 
         capture_subunit = captureSubUnit,
         capture_area = realCaptureArea) %>%
  relocate(capture_area, .before = capture_unit) %>% 
  relocate(capture_subunit, .after = capture_area) %>%
  mutate(capture_area = gsub(" ", "", capture_area)) %>%
  mutate(capture_unit = gsub(" ", "", capture_unit)) %>%
  distinct()

```


### GPS data table formating

Yet again, I pulled the columns that I wanted from the raw Wildlife Tracker data
and renamed and ordered them appropriately. This included collarID, 
datYearAndJUlian, litutde, and longitude. I then had to tell R that the 
timestamp column is a time and date, as well as the format and timezone of that 
time and date.  

```{r gps_data_data, eval = FALSE, echo = TRUE}

gps_data <- raw_data %>%
  select(collarID, dateYearAndJulian, latitude, longitude) %>%
  rename(collar_id = collarID, 
         timestamp = dateYearAndJulian) %>%
  relocate(longitude, .before = latitude) %>%
  mutate(timestamp = mdy_hm(timestamp, tz = "America/Denver")) 

```

There were then a number of steps I had to take to convert the fix points from 
lat long to utm. First, I had to turn the lat long columns into a spatial object
so that R could recognize them as such. Then I could transform that spatial object
to utm coordinates and pull just those columns from the data. I then joined those
utm easting and northing columns back with the object I created from the 
Wildlife Tracker data, deleted the lat long columns, and renamed the utm columns
so that they matched my database table. 

``` {r spatial_object, eval = FALSE, echo = TRUE}

utm_coords <- gps_data %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32612) %>%
  st_coordinates()


gps_data <- cbind(gps_data, utm_coords) %>%
  select(collar_id, timestamp, X, Y) %>%
  rename(utm_x = X, utm_y = Y)

```

## Adding Data to the Database

Now that the data is cleaned and organized according to the format of the 
monitoring techniques database tables, we can actually add the data to the tables.
This is a simple process that can be done with a single line of code for each table.

Since this is our first time adding we can merely tell it to append the data to 
the table. However, in the future when we have to be careful about whether we're 
telling it to append  or overwrite the data. In cases where we're downloading all 
the data that exists for a table we can tell it to overwrite, but in cases where 
we're merely adding data that's new since the last time we downloaded we need to 
make sure we're telling it to append that data to what already exists in the table.

### Carnivore table

Here I added the data I pulled and formatted from the Wildlife Tracker raw data to 
the carnivores table in the montech.db.

```{r carnivore_table1, eval = FALSE, echo = TRUE}

dbWriteTable(montech_db, "carnivores", carnivores, append = TRUE)

```

### Collars table

Here I added the data I pulled and formatted from the Wildlife Tracker raw data to 
the collars table in the montech.db.

```{r collar_table1, eval = FALSE, echo = TRUE}

dbWriteTable(montech_db, "collars", collars, append = TRUE)

```

### Deployments table

Here I added the data I pulled and formatted from the Wildlife Tracker raw data
to the deployments table in the montech.db.

```{r capture_sites_table, eval = FALSE, echo = TRUE}

dbWriteTable(montech_db, "deployments", deployments, append = TRUE)

```

### Capture sites table

Here I added the data I pulled and formatted from the Wildlife Tracker raw data to 
the capture sites table in the montech.db.

```{r capture_sites_table1, eval = FALSE, echo = TRUE}

dbWriteTable(montech_db, "capture_sites", capture_sites, append = TRUE)

```

### GPS data table

Here I added the data I pulled and formatted from the Wildlife Tracker raw data to 
the gps data table in the montech.db.

```{r gps_data_table1, eval = FALSE, echo = TRUE}

dbWriteTable(montech_db, "gps_data", gps_data, append = TRUE)

```

